\section{Introduction}\label{sec:intro}

\sebi{Classes of bounded expansion and} nowhere dense classes of graphs were introduced 
by Ne\v set\v ril and Ossona de 
Mendez~\cite{nevsetvril2008grad,nevsetvril2010first,nevsetvril2011nowhere} as a very 
general model
for uniform sparseness of graphs. These classes generalize many 
familiar classes of sparse graphs, such as planar graphs, graphs 
of bounded treewidth,  graphs of bounded degree, and, in fact, 
all classes that exclude a fixed 
topological minor.
Formally, a class $\CCC$ of graphs is {\em{nowhere dense}} if there is a function $t\colon \N\to \N$ such that for every $r\in \N$, no graph $G$ in~$\CCC$ contains the clique $K_{t(r)}$ on $t(r)$ vertices  as an {\em{$r$-shallow minor}},
i.e., as a subgraph of a graph obtained from $G$ by contracting mutually disjoint  subgraphs of radius $r$   to single vertices.
\sebi{Classes of bounded expansion~\cite{nevsetvril2008grad} 
are important subclasses 
of nowhere dense graph classes which are defined by bounds on 
the edge density of $r$-shallow minors.}


\sebi{Especially} the concept of nowhere denseness
turns out to be very robust, as witnessed by the fact that it is equivalent 
to multiple other concepts studied in different areas of mathematics. 
One can equivalently characterize nowhere dense graph classes 
by bounds on the density of (topological) shallow 
minors~\cite{nevsetvril2010first,nevsetvril2011nowhere},
quasi-wideness~\cite{nevsetvril2011nowhere} (a notion introduced by
Dawar~\cite{dawar2010homomorphism} in his study of homomorphism
preservation properties), low tree-depth
colorings~\cite{nevsetvril2008grad}, generalized coloring
numbers~\cite{zhu2009coloring}, sparse neighborhood
covers~\cite{GroheKRSS15,grohe2014deciding}, by a game called the
splitter game~\cite{grohe2014deciding}, and by the model-theoretic
concepts of stability and independence~\cite{adler2014interpreting}.
For a broader discussion on the graph theoretic sparsity we refer to the book
of Ne\v{s}et\v{r}il and Ossona de Mendez~\cite{sparsity}.

\begin{comment}
These alternative characterizations have been very useful in 
the design of efficient algorithms. For instance, 
the {\sc{Subgraph Isomorphism}} and {\sc{Homomorphism}} problems 
are fixed-parameter tractable on any nowhere dense
class, parameterized by the size of the pattern graph~\cite{nevsetvril2010first}
and so is the {\sc Distance-$r$ Dominating Set} problem, parameterized
by the size of the solution~\cite{DawarK09}. In fact, 
the {\sc Distance-$r$ Dominating Set} problem admits
polynomial kernels~\cite{siebertz2016polynomial} and even 
almost linear kernels on nowhere dense classes of 
graphs~\cite{eickmeyer2016neighborhood}
(see also~\cite{drange2016kernelization} for the case $r=1$). 
It was shown in~\cite{grohe2014deciding}
that every first-order definable problem can be decided in
almost linear time on any nowhere dense graph class.

It is a natural question to ask for the most general classes of graphs
which admit efficient solutions for certain problems, or to 
classify them into tractable and intractable classes. It was shown 
that for the first-order model-checking problem~\cite{dvovrak2013testing} and for
the {\sc Distance-$r$ Dominating Set} problem~\cite{drange2016kernelization} 
the dividing line for algorithmic tractability 
on subgraph closed classes of graphs is exactly between the
nowhere dense and somewhere dense graph classes. 
\end{comment}

%\bigskip

In this work we revisit the connections between the notion of nowhere 
denseness and notions from  model theory and finite model theory.
We first consider uniform quasi-wideness, a
notion  introduced by Dawar~\cite{dawar2010homomorphism}, who 
proved that every quasi-wide class that is closed under taking substructures
and disjoint unions has the homomorphism preservation property. 
Formally, a class $\CCC$ of graphs is \emph{uniformly quasi-wide} if there are
functions $N\colon \N\times\N\rightarrow \N$ and $s\colon \N\rightarrow \N$ such
that for all $r,m\in \N$ and all subsets $A\subseteq V(G)$ for
$G\in \CCC$ of size $\abs{A}\geq N(r,m)$ there is a set
$S\subseteq V(G)$ of size $\abs{S}\leq s(r)$ and a set
$B\subseteq A\setminus S$ of size $\abs{B}\geq m$ which is $r$-independent in
$G-S$. Recall that a set $B\subseteq V(G)$ is called {\em{$r$-independent}} in $G$ if all
distinct $u,v\in B$ are at distance 
larger than $r$ in $G$.
 Ne\v{s}et\v{r}il and Ossona de Mendez proved that
the notions of uniform quasi-wideness and nowhere denseness coincide for 
classes of graphs~\cite{nevsetvril2011nowhere}. 
We revisit their proof, obtaining improved bounds.


The second topic of study is the
connection between nowhere denseness and stability theory. 
Fix
% Let $\cal C$ be a class of structures over
 a relational vocabulary $\Sigma$. Let 
$\phi(\tup{x},\tup{y})$ be a $\Sigma$-formula with the free variables
partitioned into two groups $\tup{x}, \tup{y}$. A \emph{$\phi$-ladder}
of length $n$ in a $\Sigma$-structure $\str A$ is a sequence $\tup{a}_1,\ldots, \tup{a}_{n},
\tup{b}_1,\ldots, \tup{b}_{n}$ of tuples of elements of $\str A$ 
such that for all $1\leq i,j\le n$,
\[\strA\models\phi(\tup{a}_i,\tup{b}_j)\Longleftrightarrow i\leq j. \]
The least  $n$ for which 
there is no $\phi$-ladder of length $n$ is 
the \emph{ladder index} 
of $\phi(\tup{x},\tup{y})$ in $\str A$ (which may depend on the way we split the
variables). 

Based on work of Podewski and Ziegler~\cite{podewski1978stable}, 
Adler and Adler~\cite{adler2014interpreting}
proved that every nowhere dense class $\CCC$ of graphs is stable, that is, 
the ladder index of every first-order formula $\phi(\tup{x},\tup{y})$ over
graphs from $\CCC$ is bounded by a constant depending only on $\phi$ 
and~$\CCC$. In fact, for a subgraph-closed class $\CCC$, the notions of nowhere denseness and stability coincide.

We remark that the above connections between nowhere denseness and notions from model theory have recently found algorithmic applications.
Both uniform quasi-wideness and stability techniques are key tools used in the study of the complexity of the {\sc Distance-$r$ Dominating Set} problem on nowhere dense graph classes,
and in particular in the design of polynomial kernelization procedures for this problem~\cite{DawarK09,drange2016kernelization,eickmeyer2016neighborhood,siebertz2016polynomial}.

\sebi{One of the main contributions of \cite{eickmeyer2016neighborhood}
was to prove that nowhere dense classes of graphs have small
\emph{distance neighborhood complexity}. More precisely, it
was proved that for every nowhere dense class $\CCC$ of 
graphs, every positive integer $r$ and every $\epsilon>0$ there
exists a constant $c$ such that for every graph $G\in\CCC$ and
every vertex subset $A\subseteq V(G)$, the set of distance-$r$ neighborhood intersections with $A$,
$\{N_r[v]\cap A : v\in V(G)\}$,
has size bounded by $c\cdot |A|^{1+\epsilon}$. This result
generalizes a result of Reidl et al.~\cite{reidl2016characterising}, who proved that classes of bounded expansion have linear distance neighborhood complexity. }

\sebi{Distance neighborhood complexity is a 
special case of a much more general definition which is 
commonly studied in (infinite) model theory. 
Let $\psi(\tup x,y)$ be a first-order formula, where 
$\tup x$ is an $m$-tuple of variables and $y$ is a single variable. 
Let $\strA$ be a structure and let $A$ be a set of elements of
$\strA$. Then the \emph{$\psi$-type} $S_\psi(A,\strA)$ of $A$ in 
$\strA$ is the set
\[\{\{\tup a\ \in A^m : \strA\models\psi(\tup a,b)\} : b\in V(\strA)\}.\]
In stability theory, the number of types that can be realized over a parameter
set $A$ in a structure~$\strA$ of a theory $T$ 
is one of the key measures for the complexity of the theory $T$, 
in fact, the original definition of a stable theory is made in 
terms of bounds on the number of types that are realized in
its models. }

\paragraph{Our contribution.} 
Our first result is  a new proof of a result of
Ne\v{s}et\v{r}il and Ossona de Mendez~\cite{nevsetvril2010first},
which states that a class $\CCC$ of graphs is nowhere dense if and only if it
is uniformly quasi-wide. The proof of Ne\v{s}et\v{r}il 
and Ossona de Mendez goes back to a construction
of Kreidler and Seese~\cite{kreidler1998monadic} (see also Atserias et al.~\cite{atserias2006preservation}), 
and uses iterated Ramsey arguments. Hence the original bounds on 
the function $N$ are huge. Recently, Kreutzer, Rabinovich and the second author
 proved that we may always choose~$N$ to be a polynomial 
function of $m$, with the degree of the polynomial depending on $r$~\cite{siebertz2016polynomial}. However, the exact dependence of degree of the polynomial on $r$ and on the class $\CCC$ itself
 was  not specified in~\cite{siebertz2016polynomial}, as the existence of a polynomial bound is derived
from non-constructive arguments used by Adler and Adler's in showing that every nowhere dense class of graphs
is stable~\cite{adler2014interpreting}. We give a new construction 
which is considerably simpler than that of~\cite{siebertz2016polynomial}
and which gives explicit bounds on the degree of the polynomial. 
More precisely, we prove the following theorem; here, the notation $\Oof_{r,t}(\cdot)$ hides computable factors depending on $r$ and $t$.

\begin{theorem}\label{thm:new-uqw}
For all $r,t\in \N$ there is a polynomial  $N\colon \N\to \N$ with $N(m)=
\Oof_{r,t}{(m^{{(2t+1)}^{2rt}})}$, such that the following holds.
Let $G$ be a graph such that $K_t\not\minor_{\lfloor 5r/2\rfloor} G$, and
let $A\subseteq V(G)$ be a vertex subset of size at least $N(m)$, for a given $m$.
Then there exists a set $S\subseteq V(G)$ of size $|S|\leq t$ and a set $B\subseteq A\setminus S$ 
of size $|B|\geq m$ which is $r$-independent in $G-S$.
Moreover, given~$G$ and $A$, such sets $S$ and $B$ can be computed in time $\Oof_{r,t}(|A|\cdot |E(G)|)$. 
\end{theorem}

Let us remark
that even though the techniques employed to prove \cref{thm:new-uqw} are inspired by methods from stability theory, 
at the end we use only very simple graph theoretic notions. In particular, as asserted in the last sentence of the theorem, the
proof can easily be turned into an efficient algorithm.

 Podewski and Ziegler~\cite{podewski1978stable} 
consider \emph{flat graphs}, a notion corresponding to uniform quasi-wideness in the 
infinite. They show that  flat graphs are stable using an 
infinite Ramsey argument and compactness. The observation of Adler and Adler~\cite{adler2014interpreting} was that
this can be translated to classes $\CCC$ of finite graphs by considering an infinite graph that is the disjoint union of graphs from~$\CCC$.
However, this inherently non-constructive approach cannot give explicit upper bounds on parameters governing the stability of $\CCC$, for instance, the ladder indices.

Based on the approach of Podewski and Ziegler~\cite{podewski1978stable}, we give a combinatorial 
proof that every first-order formula has finite ladder index on every
nowhere dense class, which does not involve infinite combinatorics and model theory.
In particular, instead of compactness we use Gaifman's Locality Theorem for
first-order logic~\cite{gaifman1982local}. The following theorem summarizes our~result.

\begin{theorem}\label{thm:new-stable}
  There are computable functions $f\colon \N^3\to\N$ and $g\colon\N\to\N$ with the following property.
Suppose $\phi(\bar x,\bar y)$ is a formula of quantifier rank $q$ and with $d$ free variables,
and $G$ is a graph such that $K_t\not\minor_{g(q)} G$. Then the ladder index of $\phi(\bar x,\bar y)$ in $G$ is at most $f(q,d,t)$.
\end{theorem}

Note that in particular, \cref{thm:new-stable} implies that every nowhere dense graph is stable, which was the main conclusion of Adler and Adler~\cite{adler2014interpreting}.

\bigskip
Finally, we prove the following theorem concerning the number of
types in sparse graph classes. 
\begin{change}{sz}For a graph $G$, a set of vertices $A\subset V(G)$, and a formula $\psi(\tup x, y)$, define the \emph{set of $\psi$-types} over $A$:
\[S_\psi(A,G)=\setof{\setof{\tup a\ \in A^m}{G\models\psi(\tup a,v)}}{ v\in V(G)}\]
\end{change}


\begin{theorem}\label{thm:vc-density}
Let $\CCC$ be a class of graphs and let $\psi(\tup x,y)$ be a first-order formula, where 
$\tup x$ is an $m$-tuple of variables and $y$ is a single variable. 
\begin{enumerate}
\item If $\CCC$ is nowhere dense, then for every $\epsilon>0$ 
there exists a constant~$c$ such that for every $G\in \CCC$ and every
$A\subseteq V(G)$, we have $|S_\psi(A,G)|\leq c\cdot |A|^{1+\epsilon}.$

\item If $\CCC$ has bounded expansion, then there exists a constant~$c$ such that for every $G\in \CCC$ and every $A\subseteq V(G)$, we have $|S_\psi(A,G)|\leq c\cdot |A|$.
\end{enumerate}
\end{theorem}
\sebi{The proof of \Cref{thm:vc-density} is (again) based on Gaifman's locality
theorem for first-order logic, which allows us to focus at local 
neighborhoods around the elements $v\in V(G)$. 
Using the closure lemma of~\cite{DrangeDFKLPPRVS16} for
classes of bounded expansion and its generalization to
nowhere dense classes~\cite{eickmeyer2016neighborhood}
together with uniform quasi-wideness,
we are then able to well control and classify 
the interaction of elements $v$ with the set $A$. }




\begin{change}{sz}
\Cref{thm:vc-density} shows a strong new connection 
	between first-order logic and the notions of bounded expansion
	and nowhere denseness.
	It is not difficult to see that it also yields new dichotomies 
characterizing classes which are nowhere dense  or have bounded expansion, obtained by combining~\cref{thm:vc-density}
with the complementary result stated below.

\begin{theorem}\label{thm:vc-density-lower-bound}
Let $\CCC$ be a class of graphs which 
is closed under taking subgraphs. 
\begin{enumerate}
\item If $\CCC$ is somewhere dense, then there is a formula 
$\psi(x,y)$ such that for every $n\in \N$ there are $G\in\CCC$ and $A\subseteq V(G)$ 
with $|A|\geq n$ and $|S_\psi(A,G)|=2^{|A|}$. 
\item If $\CCC$ has unbounded expansion, then there is a formula 
$\psi(x,y)$ such that for every function $f:\N\rightarrow \N$ 
there are $G\in\CCC$ and $A\subseteq V(G)$ 
with $|S_\psi(A,G)|>f(|\psi|)\cdot |A|$. 
\end{enumerate}
\end{theorem}
\end{change}



\pagebreak


% Finally, we observe that an argument of Bousquet and
% Thomasse\'e~\cite{BousquetT15} can be slightly modified to prove that
% the VC-dimension of the $r$th power $G^r$ of a graph $G$
% with $K_t\not\minor_r G$ is bounded by $t-1$.
% Boundedness of VC-dimension is a weaker condition than the stability of a class, however it is sufficient in many contexts.
% Therefore, we consider investigating explicit bounds on the VC-dimension potentially useful for future research.
%
% \begin{theorem}\label{thm:new-vc}
% Let $G$ be a graph such that $K_t\not\minor_r G$. Then the
% VC-dimension of $G^r$, the $r$th power of the graph~$G$, is bounded by $t-1$.
% \end{theorem}

%\paragraph{Organization.} We give background from graph theory in \cref{sec:uqw}, where we also
%prove \cref{thm:new-uqw}. 
%We provide background 
%on logic and prove \cref{thm:new-stable} in \cref{sec:stable}. 



