\section{Preliminaries}\label{sec:prelim}
We  recall some basic notions from graph theory. 

\medskip
All graphs in this paper are finite, undirected and simple, that is, 
they do not have loops or parallel edges. Our notation is standard,
we refer to~\cite{diestel2012graph} for more background on 
graph theory. 
We write $V(G)$ for the vertex set of a graph $G$ and
$E(G)$ for its edge set. 
The {\em{distance}} between vertices $u$ and $v$ in $G$, denoted $\dist_G(u,v)$, is the length of a shortest path between $u$ and $v$ in~$G$.
If there is no path between $u$ and $v$ in $G$, we put $\dist_G(u,v)=\infty$.
The {\em{(open) neighborhood}} of a vertex $u$, denoted $N(u)$, is the set of neighbors of $u$, excluding $u$ itself.
For a non-negative integer $r$, by $N_r[u]$ we denote the {\em{(closed) $r$-neighborhood}} of $u$ which comprises vertices at distance at most $r$ from $u$; 
note that $u$ is always contained in its closed $r$-neighborhood. The \emph{radius} of a connected graph $G$ is the least integer $r$ such that there is some vertex $v$ of $G$ with $N_r[v]=V(G)$.


A {\em{minor model}} of a graph $H$ in $G$ is a family $(I_u)_{u\in V(H)}$ of pairwise vertex-disjoint connected subgraphs of $G$, called {\em{branch sets}},
such that whenever $uv$ is an edge in~$H$, there are $u'\in V(I_u)$ and $v'\in V(I_v)$ for which $u'v'$ 
is an edge in $G$.
The graph $H$ is a {\em{depth-$r$ minor}} of $G$, denoted $H\minor_rG$, if there is a minor model
$(I_u)_{u\in V(H)}$ of~$H$ in $G$ such that each $I_u$ has radius at most $r$.

A class $\CCC$ of graphs is \emph{nowhere dense} if there is a function 
$t\colon \N\rightarrow \N$ such that for all $r\in \N$ it holds that $K_{t(r)}\not\minor_r G$
for all $G\in \CCC$, where $K_{t(r)}$ denotes the clique on $t(r)$ vertices.
The class~$\CCC$ moreover has \emph{bounded expansion}
if there is a function $d\colon\N\rightarrow\N$ such that for all 
$r\in \N$ and all $H\minor_rG$ with $G\in\CCC$, the {\em{edge density}}
of $H$, i.e. $|E(H)|/|V(H)|$, is bounded by $d(r)$. Note that every 
class of bounded expansion is nowhere dense. The converse is not necessarily true in general~\cite{sparsity}.

% A set $B\subseteq V(G)$ is called {\em{$r$-independent}} in a graph $G$ if  $\dist_G(u,v)>r$ for all
% distinct $u,v\in B$.

 

A set $B\subseteq V(G)$ is called {\em{$r$-independent}} in a graph $G$ if  $\dist_G(u,v)>r$ for all
distinct $u,v\in B$. 
A class $\CCC$ of graphs is \emph{uniformly quasi-wide} if for every $r\in \N$ there is a number $s\in \N$
and a function $N\from\N\to\N$ such that
for every $m\in \N$, graph $G\in \CCC$, and vertex subset $A\subseteq V(G)$ of size $\abs{A}\geq N(m)$, there is a set
$S\subseteq V(G)$ of size $\abs{S}\leq s$ and a set
$B\subseteq A-S$ of size $\abs{B}\geq m$ 
which is $r$-independent in $G-S$.
Recall that Ne\v set\v ril and Ossona de 
Mendez proved~\cite{nevsetvril2011nowhere} that nowhere dense graph classes are exactly the same as uniformly quasi-wide classes. 
The following result of Kreutzer, Rabinovich and the second author~\cite{siebertz2016polynomial}
improves their result, by showing that the function $N$ can be taken polynomial:

\begin{theorem}[\cite{siebertz2016polynomial}]\label{thm:krs}
For every nowhere dense class $\CCC$
and for all $r\in \N$ there is a polynomial $N\from \N\to\N$ 
and a number $s\in \N$ such that the following holds.
Let be $G\in \CCC$ be aa graph and let $A\subset V(G)$ be a vertex subset of size at least $N(m)$, for a given $m$.
Then there exists a set $S\subset V(G)$ of size $|S|<s$
and a set $B\subset A-S$ of size $|B|\ge m$ which is $r$-independent in $G-S$.	
\end{theorem}

As we mentioned, the proof of Kreutzer et al.~\cite{siebertz2016polynomial} relies on non-constructive arguments and does not yield explicit bounds on $s$ and (the degree of) $N$. 
In the next section, we discuss a further strengthening of this result, by providing explicit bounds on the function $N$ and the number $s$.